---
title: "Multi-state model for prognosis of hypertension"
format: html
author: Yiling Zhou
R version: 4.2.3
Input data: rx_hyper_med, rx_adhd_med, cohort_admin, dx_outcomes, dx_adhd_diag
---

# Load R packages

To avoide the functions from data.table are masked by those with same names from other packages, please load data.table lastly.

tidyverse and dplyr are required, but do not load them here; instead, load them at the specified location by using tidyverse::fun.

```{r}
library(tableone)
library(ggplot2)
library(IRanges)
library(survminer)
library(survival)
library(tibble)
library(data.table)
```

# Change the working directory

```{r}
knitr::opts_chunk$set(root.dir = "H:/Project/Multistate_ADHDandHTN")
```

# Provide general information

**recruit_start**: the start date from which study population was selected. 
**recruit_end**: the end date from which study population was selected 2020-12-31
**admin_end**: the end date of the follow up: 2022-12-31.

```{r}
recruit_start <- as.IDate("2013-01-01")
recruit_end <- as.IDate("2020-12-31")
admin_end <- as.IDate("2022-12-31")
```

# Import data

```{r}
rx_hyper_med <- fread("processData/rx_hyper_med.csv", header = TRUE, sep = "," ,colClasses = list(character = "pin"))

cohort_admin <- fread("processData/cohort_admin.csv", header = TRUE, sep = ",", colClasses = list(character = "pin"))

rx_adhd_med <- fread("processData/rx_adhd_med.csv", header = TRUE, sep = ",", colClasses = list(character = "pin"))
dx_adhd_diag <- fread("processData/dx_adhd_diag.csv", header = TRUE, sep = ",", colClasses = list(character = "pin"))

dx_outcomes <- fread("processData/dx_outcomes.csv", header = TRUE, sep = "," ,colClasses = list(character = "pin"))
```

# Data clean

## Format all variables

**Assumption**: dispensation at the same date was considering as one combined treatment regimen, with the supply from all the dispensations comprising a single treatment duration.

```{r}
## cohort_admin
date_names <- names(cohort_admin)[names(cohort_admin) %like% "_date$"]
cohort_admin[, (date_names) := lapply(.SD, as.IDate), .SDcols = date_names]
cohort_admin[, (setdiff(names(cohort_admin), c("pin", date_names, "SES"))) := lapply(.SD, as.integer), .SDcols = setdiff(names(cohort_admin), c("pin", date_names, "SES"))]
cohort_admin <- unique(cohort_admin)
setnames(cohort_admin, old="id",new="intellectual")

## rx_adhd_med
rx_adhd_med <- rx_adhd_med[, adhd_disp_date := as.IDate(adhd_disp_date)][!is.na(adhd_disp_date),]

## dx_adhd_diag
dx_adhd_diag <- dx_adhd_diag[, adhd_diag_date := as.IDate(adhd_diag_date)][!is.na(adhd_diag_date),]
dx_adhd_diag <- unique(dx_adhd_diag)

## rx_hyper_med
rx_hyper_med <- rx_hyper_med[, disp_date := as.IDate(disp_date)][!is.na(disp_date),]
rx_hyper_med <- unique(rx_hyper_med)

## dx_outcomes
date_names <- names(dx_outcomes)[names(dx_outcomes) %like% "_date$"]
dx_outcomes[, (date_names) := lapply(.SD, as.IDate), .SDcols = date_names]
```

## Deal with multiple prescriptions for ADHD, hypertension on the same date

**Assumption**: multiple prescriptions filled on the same day are considered as one combined treatment regimen, with the supply from all the prescriptions comprising a single treatment duration.

```{r}
rx_adhd_med <- rx_adhd_med[, .SD[1], by =.(pin, adhd_disp_date)]
rx_hyper_med <- rx_hyper_med[, .SD[1], by =.(pin, disp_date)]
```

# Part 1: Identify study population (P1)

One output in this part: P1_Participants_excluded.csv

## 1. identify the individuals newly diagnosed with hypertension

### Step 1: identify individuals starting C03A, C03B C08C, C09 between recruit_start and recruit_end, e.g., 2020-12-31 (excl. recruit_end)

hyper_patient is data.table, with two variables, pin, their baseline date. One row per individual.

```{r}
hyper_patient <- rx_hyper_med[atc %in% c("C03A", "C03B", "C08C", "C08G", "C09A", "C09B", "C09C", "C09D"),][disp_date>=recruit_start & disp_date<recruit_end, ] 
hyper_patient[, baseline_date := min(disp_date, na.rm = TRUE), by = pin] 
hyper_patient <- unique(hyper_patient[, .(pin, baseline_date)])
```

### Step 2: had not been prescribed any these medication C03, C08, C09 before baseline (excl.baseline)

```{r}
exclude_patient <- unique(rx_hyper_med[pin%in%hyper_patient$pin,])[hyper_patient, on = .(pin = pin)][disp_date < baseline_date,]
hyper_patient1 <- hyper_patient[!pin%in%exclude_patient$pin, ]
rm(exclude_patient)
```

### Step 3: these medications were not used to treat other diseases at baseline, without established CHD, without CKD before or at baseline

i.e., without any diagnosis of CKD, ESKD, diabetic nephropathy, glomerular diseases, HF, CHD, or without medication that are served as proxy for MACE and CKD, before and at baseline: otherIndication == 0

```{r}
hyper_patient1 <- cohort_admin[hyper_patient1, on = .(pin = pin)]
hyper_patient2 <- copy(hyper_patient1)
```

### 2. apply exclusion criteria

```{r}
# Create baseline_mn and baseline_yr to prepare for calculating baseline age
hyper_patient2[, `:=` (baseline_mn = data.table::month(baseline_date), baseline_yr =  data.table::year(baseline_date))]
# Calculate the baseline age
hyper_patient2[(birth_mn > baseline_mn), age :=  baseline_yr - birth_yr - 1]
hyper_patient2[(birth_mn <= baseline_mn), age :=  baseline_yr - birth_yr]

# 1. exclude individuals with missing data in sex, birth year, birth month
hyper_patient3 <- hyper_patient2[!is.na(sex), ][!is.na(birth_yr),][!is.na(birth_mn),]

# 2. exclude age < 18 years
hyper_patient4 <- hyper_patient3[age>=18&age<=90, ]

# 3. exclude who emigrated before or at baseline 
hyper_patient5 <- hyper_patient4[is.na(emigration_date)|emigration_date > baseline_date,]

# 4. exclude individuals with error data, i.e.,who died before or at baseline 
hyper_patient6 <- hyper_patient5[is.na(death_date)|death_date > baseline_date,]

# 5. exclude who had a outcome, HF, MI, HF, CR death, non-CR death, unnatural death before baseline or at baseline
hyper_patient6 <- dx_outcomes[hyper_patient6, on = .(pin = pin)]
hyper_patient7 <- hyper_patient6[is.na(stroke_date)|stroke_date > baseline_date,][is.na(AMI_date)|AMI_date > baseline_date,][is.na(HHF_date)|HHF_date > baseline_date,][is.na(CKD_date)|CKD_date > baseline_date,][is.na(unnaturaldeath_date)|unnaturaldeath_date > baseline_date,][is.na(CRdeath_date)|CRdeath_date > baseline_date,][is.na(noCRdeath_date)|noCRdeath_date > baseline_date,][otherIndication == 0,]

## the final study population, including all variables in cohort admin 
dt <- copy(hyper_patient7)
## remove useless columns
dt[, baseline_mn:=NULL]
dt[, birth_yr:=NULL]
dt[, birth_mn:=NULL]
dt[, otherIndication:=NULL]
```

#### P1_Participants_excluded.csv

Export table for flowchart

```{r}
Participants_excluded <- tibble::tibble(all = uniqueN(hyper_patient$pin), 
                                washout = uniqueN(hyper_patient1$pin), 
                                otherIndications = uniqueN(hyper_patient2$pin),
                                missing = uniqueN(hyper_patient3$pin),
                                age = uniqueN(hyper_patient4$pin), 
                                emigration = uniqueN(hyper_patient5$pin), 
                                die = uniqueN(hyper_patient6$pin), 
                                outcome = uniqueN(hyper_patient7$pin))

write.csv2(Participants_excluded, file = "results/P1_participants_excluded.csv", quote = F)
```

remove useless data table/data frame/tibble

```{r}
delete <- ls(pattern = "^hyper_patient")
rm(list = c(delete, "Participants_excluded", "cohort_admin", "dx_outcomes"))
gc()
```

# Define variables

## 1. define key variables

**age_cat**: age group at baseline, young, middle age, elderly.

**baseline_yr**: calendar year of baseline, already defined at the prior step-age calculation.

**sex**: 0 is female, 1 is male.

**follow_end**: date, end of follow up for each individual, the earliest date among emigration_date, death_date, admin_end.

Other variaibles have been defined during the preparation for cohort admin.

```{r}
dt[, age_cat := cut(age, breaks = c(17, 34, 64, Inf), labels = c("young adult", "middle-aged", "elderly"), right = TRUE)]
dt[, baseline_yr_cat := fifelse(baseline_yr <2018, 0, 1)]
dt[, sex:=factor(sex, levels = c(1, 0), labels = c("male", "female"))]

dt[, follow_end := pmin(emigration_date, death_date, admin_end, na.rm = T)]

## remove useless columns
dt[, emigration_date:=NULL]
```

## 2. define baseline ADHD status (diagnosis and medication)

**adhd_baseline**: baseline ADHD status (0 vs 1), assessed using data from start of registers to baseline.

**adhd_date**: date, the first date of ADHD diagnosis based on diagnosis and medication.

```{r}
## first date of adhd based on diagnosis and medication
adhd <- rx_adhd_med[,.(pin, adhd_disp_date)]
setnames(adhd, old = "adhd_disp_date", new = "adhd_diag_date")
adhd <- unique(rbind(adhd, dx_adhd_diag))
adhd[, adhd_date := min(adhd_diag_date, na.rm = TRUE), by = pin]
adhd <- unique(adhd[,.(pin, adhd_date)])

### join with dt and define baseline ADHD status
dt <- adhd[dt, on=.(pin = pin)]
dt[, adhd_baseline:=fifelse(is.na(adhd_date)|adhd_date>baseline_date, 0, 1)]
dt[, adhd_baseline:=factor(adhd_baseline, levels = c(0, 1), labels = c("no ADHD", "ADHD"))]

# remove data
rm(adhd)
rm(dx_adhd_diag)
```

## 3. define outcome and survival time 

**CRdeath** (0 vs 1).

**CRdeath_time**: survival time.

Note:
1. dt_outcomes: if AMI, stroke, HHF, or CKD happens on the same date of death, then CR event will be as 1 and death will happen after 1 day (fatal event).
2. adjust the CV and CKD dates happen on the same date: only 4 individuals

```{r}
# CR death
dt[, CRdeath_date:=fifelse(CRdeath_date>follow_end, as.IDate(NA), CRdeath_date)]
dt[, CRdeath := fifelse(is.na(CRdeath_date)|CRdeath_date>follow_end, 0, 1)]
dt[CRdeath == 1, CRdeath_time := CRdeath_date - baseline_date]
dt[CRdeath == 0, CRdeath_time := follow_end - baseline_date]

# non-CR death and unnatural death
dt[, noCRdeath_date:=pmin(noCRdeath_date, unnaturaldeath_date, na.rm = TRUE)]
dt[, unnaturaldeath_date:=NULL]
dt[, noCRdeath_date:=fifelse(noCRdeath_date>follow_end, as.IDate(NA), noCRdeath_date)]
dt[, noCRdeath := fifelse(is.na(noCRdeath_date)|noCRdeath_date>follow_end, 0, 1)]
dt[noCRdeath == 1, noCRdeath_time := noCRdeath_date - baseline_date]
dt[noCRdeath == 0, noCRdeath_time := follow_end - baseline_date]

# HHF
dt[, HHF_date:=fifelse(HHF_date>follow_end, as.IDate(NA), HHF_date)]
dt[, HHF := fifelse(is.na(HHF_date)|HHF_date>follow_end, 0, 1)]
dt[HHF == 1, HHF_time := HHF_date - baseline_date]
dt[HHF == 0, HHF_time := follow_end - baseline_date]

# AMI
dt[, AMI_date:=fifelse(AMI_date>follow_end, as.IDate(NA), AMI_date)]
dt[, AMI := fifelse(is.na(AMI_date)|AMI_date>follow_end, 0, 1)]
dt[(!is.na(AMI_date) & HHF_date==AMI_date), AMI_date:= AMI_date-1]
dt[AMI == 1, AMI_time := AMI_date - baseline_date]
dt[AMI == 0, AMI_time := follow_end - baseline_date]


# stroke
dt[, stroke_date := fifelse(stroke_date>follow_end, as.IDate(NA), stroke_date)]
dt[, stroke := fifelse(is.na(stroke_date)|stroke_date>follow_end, 0, 1)]
dt[(!is.na(stroke_date) & HHF_date==stroke_date), stroke_date:= stroke_date-1]
dt[(!is.na(stroke_date) & AMI_date==stroke_date), stroke_date:= stroke_date-1]
dt[stroke == 1, stroke_time := stroke_date - baseline_date]
dt[stroke == 0, stroke_time := follow_end - baseline_date]

# CKD
dt[, CKD_date:=fifelse(CKD_date>follow_end, as.IDate(NA), CKD_date)]
dt[, CKD := fifelse(is.na(CKD_date)|CKD_date>follow_end, 0, 1)]
dt[(!is.na(CKD_date) & CKD_date==AMI_date), CKD_date:= CKD_date-1]
dt[(!is.na(CKD_date) & CKD_date==stroke_date), CKD_date:= CKD_date-1]
dt[(!is.na(CKD_date) & CKD_date==HHF_date), CKD_date:= CKD_date-1] 
## adjust for total 4 individuals
dt[CKD == 1, CKD_time := CKD_date - baseline_date]
dt[CKD == 0, CKD_time := follow_end - baseline_date]
```

# Part 2: Descriptive analysis (P2)

## S1. descriptive analysis for baseline table: baseline ADHD vs no ADHD

*P2_S1_descriptives_result.csv*

**follow_end_days**: follow end date - baseline date

```{r}
dt[, ESKD_date:=fifelse(ESKD_date>follow_end, as.IDate(NA), ESKD_date)]
dt[, ESKD := fifelse(is.na(ESKD_date)|ESKD_date>follow_end, 0, 1)]
dt[, admission_follow := admin_end- baseline_date]
dt[, follow_end_days := follow_end - baseline_date]
dt[, psych_number := asd+sud+scz+dep+anx+bip+ed+pd+intellectual]
dt[, psych_number := fifelse(psych_number>2, 3, psych_number)] # levels: 0, 1, 2, 3 or over
dt[, psych_any := fifelse(psych_number == 0, 0, 1)]
dt[, nd:= pmax(asd,intellectual)]
dt[, ma:= pmax(anx, dep, bip)]
dt[, psych_med_any := fifelse(add_med+anxio_med+dep_med+ep_med+hy_se_med+psycho_med == 0, 0, 1)]
dt[, age_sex_group:=paste0(age_cat, "_", sex)]

vars <- c("age", "age_cat", "sex", "age_sex_group","SES", "admission_follow", "follow_end_days", "psych_any", "nd", "asd", "intellectual", "ma", "anx",  "bip", "dep", "scz", "pd", "ed", "sud", "psych_med_any","add_med", "anxio_med", "dep_med","ep_med", "hy_se_med", "psycho_med", "t2d", "hyperlipid")
catVars<-c("age_cat", "sex", "age_sex_group","SES", "psych_any", "nd", "asd", "intellectual", "ma", "anx",  "bip", "dep", "scz", "pd", "ed", "sud", "psych_med_any", "add_med", "anxio_med", "dep_med","ep_med", "hy_se_med", "psycho_med")
nonnormals <- c("age","admission_follow", "follow_end_days")

table <- CreateTableOne(vars = vars,
                        strata = "adhd_baseline",
                        data = dt,
                        factorVars = catVars,
                        addOverall = TRUE)
P2_S1_result <- print(table,
                   showAlLevels = TRUE,
                   nonnormal = nonnormals,
                   formatOptions = list(big.mark = ","),
                   pDigits =3,
                   catDigits =1,
                   conDigists =1,
                   quote=FALSE,
                   noSpace = TRUE,
                   printToggle = FALSE,
                   addOverall =TRUE)

write.csv2(P2_S1_result, file = "results/P2_S1_descriptives_result.csv",quote = F)
rm(table)
rm(P2_S1_result)
```

# Define time-varying antihypertensive medication on or off

## Step 1: update rx_hyper_med for study population between baseline_date to follow_end.

Note: baseline_date is one disp_date in rx_hyper_med, thus we can filter rows where disp_date \>= baseline_date. but for ADHD medication data, we do not apply this filter, since prior dispensation may last beyond the baseline date.

**disp_end**: date, the date of last dispensation of antihypertensive medication before the end of follow up.

```{r}
rx_hyper_med <- rx_hyper_med[pin %in% dt$pin, ]
rx_hyper_med[, atc := NULL]
rx_hyper_med <- dt[,.(pin, baseline_date, follow_end, age_cat, sex)][rx_hyper_med, on =.(pin=pin)]
rx_hyper_med <- rx_hyper_med[disp_date<=follow_end & disp_date>=baseline_date, ]
rx_hyper_med[, disp_end := max(disp_date, na.rm = TRUE), by = pin]
```

## Step 2: estimate the refill gap for antihypertensive medication.

**refill_gap**: days, the refill gap is defined as the days supplied per dispensation; = 90% percentile of dispensation intervals stratified by sex and age group.

**interval**: days of dispensation interval.

**hyper_each_end**: date, the end date of supply for this specified dispensation of antihypertensive medication.

```{r}
## intervals of two consecuitive dispensations
rx_hyper_med <- rx_hyper_med[order(pin, disp_date),]
rx_hyper_med[, interval :=  data.table::shift(disp_date, type = "lead") - disp_date, by = pin]
rx_hyper_med[is.na(interval), interval:= follow_end - disp_date]

## 90% of qunatiles using (interval>0 and not the last interval)
quantile_hyper_med <- rx_hyper_med[disp_date < disp_end & interval>0, ][, .(refill_gap = as.integer(quantile(interval, 0.75, na.rm =T))), by = .(age_cat, sex)]

## refill gap for each individual
rx_hyper_med <- quantile_hyper_med[rx_hyper_med, on =.(age_cat, sex)]

## hyper_each_end for each dispensation
rx_hyper_med[, hyper_each_end := disp_date + refill_gap]

## adjust the end date if they are later than follow_end.
rx_hyper_med[, hyper_each_end:= fifelse(follow_end < hyper_each_end, follow_end, hyper_each_end)]

write.csv2(quantile_hyper_med, file = "results/quantile_hyper_med.csv",quote = F)
rm(quantile_hyper_med)
```

## Step 3: identify the dates on which anti-hypertensive medication administered.

reduce the overlapping dispensation records to non-overlap treat period data.

**hyper_start**: date, start of each treated period.

**hyper_end**: date, end of that treated period.

```{r}
rx_hyper <- rx_hyper_med[, {
  ir <- reduce(IRanges(as.numeric(disp_date), as.numeric(hyper_each_end)))
  .(start = as.IDate(start(ir)), end = as.IDate(end(ir)))}, by = pin] 
setnames(rx_hyper, old = "start", new = "hyper_start")
setnames(rx_hyper, old = "end", new = "hyper_end")
```

## Step 4: identify times since baseline when on and off hypertension treatment.

**hyper_treat_date:** date, the dates on which each hypertension drug treatment period starts or ends.

**hyper_treat_type**: 2 levels: start or end, the start or end of each treat period.

**since_baseline_time**: days, the days between hyper_treat_date to baseline date.

**off_hyper_med**: 0 or 1, 1-times since baseline when without using antihypertensive medication.

**rx_hyper**: a data.table for ADHD and outcome analysis.

**rx_hyper_adhdPopulation**: data.table for ADHD medication and outcome analysis.

```{r}
## dates when start or end.
rx_hyper <- melt(rx_hyper, id.vars = "pin", measure.vars = c("hyper_start", "hyper_end"), variable.name = "hyper_treat_type", value.name = "hyper_treat_date")

rx_hyper <- rx_hyper[dt[,.(pin,baseline_date,follow_end)], on = .(pin)] 

## time since baseline when starts or ends .
rx_hyper[, since_baseline_time := hyper_treat_date - baseline_date]

## code the hypertension medication treat off and on under time since baseline
rx_hyper[, off_hyper_med := fifelse(hyper_treat_type == "hyper_end", 1, 0)]
rx_hyper <- rx_hyper[,.(pin, off_hyper_med, since_baseline_time)]

rm(rx_hyper_med)
```

# Define time-varing ADHD medication on or off

## Step 1: update rx_adhd_med for study population before or at follow_end.

**adhd_disp_end**: the last prescription date before follow_end.

```{r}
rx_adhd_med <- rx_adhd_med[pin %in% dt$pin,]
# join dt_med and rx_adhd_med
rx_adhd_med <- dt[rx_adhd_med, on = .(pin)]
rx_adhd_med <- rx_adhd_med[adhd_disp_date<=follow_end, ]
rx_adhd_med[, adhd_disp_end :=  max(adhd_disp_date, na.rm = TRUE), by = pin]
rx_adhd_med <- rx_adhd_med[order(pin, adhd_disp_date),]
```

## Step 2: estimate the refill gap of ADHD medication.

**adhd_refill_gap**: days, the refill gap is defined as the days supplied per dispensation; = 90% percentile of dispensation intervals stratified by sex and age group.

**interval**: days of dispensation interval.

**adhd_each_end**: date, the end date of supply for this specified dispensation of ADHD medication.

```{r}
## interval of two consecuitive dispensations
rx_adhd_med[, interval :=  data.table::shift(adhd_disp_date, type = "lead") - adhd_disp_date, by = pin]

rx_adhd_med[is.na(interval), interval:= follow_end-adhd_disp_date]

## adhd_refill_gap: 90% quantile of intervals (>0 and not the last dispensation).
quantile_adhd_med <- rx_adhd_med[adhd_disp_date < adhd_disp_end & interval>0 & adhd_disp_date>=baseline_date, ][, .(adhd_refill_gap = as.integer(quantile(interval, 0.75, na.rm =T))), by = .(age_cat, sex)]

rx_adhd_med <- quantile_adhd_med[rx_adhd_med, on =.(age_cat, sex)]

## adhd_each_end for each dispensation
rx_adhd_med[, adhd_each_end := adhd_disp_date + adhd_refill_gap]

## adjust the end date if they are later than follow_end.
rx_adhd_med[, adhd_each_end:= fifelse(follow_end < adhd_each_end, follow_end, adhd_each_end)]

write.csv2(quantile_adhd_med, file = "results/quantile_adhd_med.csv",quote = F)
rm(quantile_adhd_med)
```

## Step 3: identify the dates on which ADHD medication administered.

reduce the overlapping dispensation records to non-overlap treat period data.

**adhd_start**: date, start of each treated period.

**adhd_end**: date, end of that treated period.

```{r}
rx_adhd_med <- rx_adhd_med[order(pin, adhd_disp_date),]

## reduce the overlapping treated time, create variables: adhd_start and adhd_end.
rx_adhd <- rx_adhd_med[, {
  ir <- IRanges::reduce(IRanges(as.numeric(adhd_disp_date), as.numeric(adhd_each_end)))
  .(start = as.IDate(start(ir)), end = as.IDate(end(ir)))}, by = pin]  

setnames(rx_adhd, old = "start", new = "adhd_start")
setnames(rx_adhd, old = "end", new = "adhd_end")
rm(rx_adhd_med)

## select ADHD medication dispensation data from baseline date to follow-end
rx_adhd <- dt[,.(pin, baseline_date)][rx_adhd, on =.(pin)]
rx_adhd <- rx_adhd[adhd_start >= baseline_date|(adhd_start<baseline_date & adhd_end>baseline_date),]
## re-define the adhd_start if it is before baseline date
rx_adhd[, adhd_start:= fifelse(adhd_start<=baseline_date & adhd_end > baseline_date, baseline_date, adhd_start)]
rx_adhd[, baseline_date:=NULL]
```

## Step 4: identify times since baseline when ADHD treatment on and off.

**adhd_treat_date:** date, the dates on which each ADHD treat period starts or ends; NA for those never treated with ADHD medication since baseline.

**adhd_treat_type**: 3 levels: start or end or never, the start or end of each ADHD treat period, or never treated with ADHD medication since baseline.

**since_baseline_time**: days, the days between adhd_treat_date to baseline date.

**off_adhd_med**: 0 or 1, 1-times since baseline when without using ADHD medication.

```{r}
## dates on which start or end.
rx_adhd <- melt(rx_adhd, id.vars = "pin", measure.vars = c("adhd_start", "adhd_end"), variable.name = "adhd_treat_type", value.name = "adhd_treat_date")

rx_adhd <- rx_adhd[dt[,.(pin,baseline_date,follow_end)], on = .(pin)] 
## those not use ADHD medication since diagnose or after baseline date.
rx_adhd[is.na(adhd_treat_type), adhd_treat_type:="never"]

## time since baseline at which starts, ends or never happens.
rx_adhd[, since_baseline_time := adhd_treat_date - baseline_date]
rx_adhd[adhd_treat_type=="never", since_baseline_time:=0]

## code the ADHD medication treat off and on under time since baseline.
rx_adhd[, off_adhd_med := fifelse(adhd_treat_type == "adhd_end"|adhd_treat_type == "never", 1, 0)]
rx_adhd <- rx_adhd[,.(pin, off_adhd_med, since_baseline_time)]
```

# Define both time-varying hypertension and ADHD medication on and off over time

**ADHD**: 1-no use adhd medication; 0-use adhd medication.

**hyper**: 1-no use hypertension medication; 0-use hypertension medication.

**since_baseline_time**: days, time since baseline.

```{r}
rx_adhd[, `:=`(med_type = "ADHD", med_status = off_adhd_med)]
rx_hyper[, `:=`(med_type = "hyper", med_status = off_hyper_med)]
rx_adhd[, off_adhd_med:=NULL]
rx_hyper[, off_hyper_med:= NULL]

rx_med <- rbind(rx_adhd, rx_hyper)
rx_med <- dcast(rx_med, pin + since_baseline_time ~ med_type, value.var = "med_status", fill = NA, fun.aggregate = sum)

rx_med <- rx_med[order(pin,since_baseline_time), ]
rx_med[, ADHD := nafill(ADHD, type = "locf"), by = pin]
rx_med[, hyper := nafill(hyper, type = "locf"), by = pin]

rx_med[is.na(ADHD), ADHD:=1]
rm(rx_adhd)
rm(rx_hyper)
```

# Define ADHD status and diagnosis time

**adhd**: diagnosis before or at follow_end as 1, otherwise as 0.

**adhd_time**: diagnosis time from baseline.

```{r}
## the period between adhd first date and baseline date
dt[, adhd_time := adhd_date - baseline_date]
## define time-varying adhd status before follow_end,follow_end_days = CRdeath_time = noCRdeath_time.
dt[is.na(adhd_time), adhd := 0]
dt[!is.na(adhd_time)&adhd_time>follow_end_days, adhd := 0]
dt[is.na(adhd), adhd:=1]

## define adhd_time
dt[, adhd_time := fifelse(adhd==0|adhd_time <=0, 0, adhd_time)]
```

# Create time-varying exposure and covariate data

**exposure**: ADHD as time-varying variable, before diagnosis as 0, after diagnosis as 1.

**hypertension_med**: time-varying variable, off antihypertensive medication use (0) vs on antihypertensive medication use(1).

**on_adhd_med**: time-varying exposure, on ADHD medication (1), off ADHD medication (1).

```{r}
## only covariates, baseline and follow end date data.
dt_base <- dt[, !c("adhd","adhd_time","adhd_date", 
                   "death_date", "noCRdeath_date", "CRdeath_date", "stroke_date", "AMI_date", "HHF_date", "CKD_date",
                   "CRdeath", "CRdeath_time", "noCRdeath", "noCRdeath_time",
                   "AMI", "AMI_time", "HHF", "HHF_time", 
                   "stroke", "stroke_time", "CKD", "CKD_time"), with = FALSE]
## events
dt_base <- tmerge(dt_base, dt, id = pin,  
                  CRdeath = event(CRdeath_time, CRdeath),
                  noCRdeath = event(noCRdeath_time, noCRdeath),
                  AMI = event(AMI_time, AMI),
                  HHF = event(HHF_time, HHF),
                  stroke = event(stroke_time, stroke),
                  CKD = event(CKD_time, CKD))
## hyper_med: intermediate variable,0-use medication; 1-not use. as well as adhd_med
dt_base <- tmerge(data1 = dt_base, data2 = rx_med, id = pin, 
                  adhd_med = tdc(since_baseline_time, ADHD, 1),
                  hyper_med = tdc(since_baseline_time, hyper, 1))
## exposure: 0-no ADHD,1-ADHD
dt_base <- tmerge(data1 = dt_base, data2 = dt, id = pin, 
                  exposure = cumtdc(adhd_time, adhd, 0))
#### hypertension_med: antihypertensive medication, 0-not use, 1-use medication. 
dt_base <- dt_base %>% 
  mutate(on_adhd_med = dplyr::case_when(
    adhd_med == 0 ~ 1,
    TRUE ~ 0
  ))%>% 
  mutate(hypertension_med = dplyr::case_when(
    hyper_med == 0 ~ 1,
    TRUE ~ 0
  )) %>% 
  dplyr::select(-hyper_med, -adhd_med)

setDT(dt_base)
```


# Part 3: Multistate model (P3) - ADHD

## 1. set the transition matrix

**tmat**: transition matrix: total 19 transitions

```{r}
tmat <- matrix(NA, 7, 7)
dimnames(tmat) <- list(from = c("hypertension", "CKD", "stroke","AMI", "HHF","CRdeath", "noCRdeath"),
                       to = c("hypertension", "CKD", "stroke", "AMI", "HHF", "CRdeath", "noCRdeath"))
tmat[1,] <- c(NA,1,2,3,4,5,6)
tmat[2,] <- c(NA,NA,NA,NA,NA,7,8)
tmat[3,] <- c(NA,NA,NA,NA,NA,9,10)
tmat[4,] <- c(NA,NA,NA,NA,NA,11,12)
tmat[5,] <- c(NA,NA,NA,NA,NA,13,14)
tmat[6,] <- c(NA,NA,NA,NA,NA,NA,NA)
tmat[7,] <- c(NA,NA,NA,NA,NA,NA,NA)

print(tmat)
```

## 2. create multiple state transitions data

```{r}
library(mstate)

covs <- c("pin")
dt <- as.data.frame(dt)
## start state: state into which no transitions are possible, set NA for time and status in the msprep function. (ref: mstate)
msbmt <- msprep(time = c(NA, "CKD_time", "stroke_time", "AMI_time", "HHF_time", "CRdeath_time", "noCRdeath_time"),
                status = c(NA, "CKD", "stroke", "AMI", "HHF", "CRdeath", "noCRdeath"), 
                data = dt, 
                trans = tmat,
                keep = covs) 

events(msbmt) %>% print()

setDT(dt)
msbmt <- as_tibble(msbmt)
msbmt <- msbmt %>% dplyr::select(pin, Tstart, Tstop, time, trans, status) 
setDT(msbmt)
```

## 3. estimate cox model for each transition

### Step 1: functions

```{r}
# Define format function
myformat <- function(x, digit = 4, sep = TRUE) {format(round(as.numeric(x), digit), nsmall = digit, trim = TRUE, big.mark = ifelse(sep, ",", ""))}

# Define Cox regression function
cox_analysis <- function(data, formula, transition_number, part){
  formula_vars <- all.vars(formula)
  for (var in formula_vars) {
    if (var %in% names(data) && length(unique(data[[var]])) == 1) {
      formula <- update(formula, as.formula(paste(". ~ . -", var)))}} 
  # Automatically remove variables in formulas with only one unique value. id = pin can be omit for two-state model, since it is automatically set.
  model <- coxph(formula, data = data, id = pin, model = TRUE) # robust = TRUE and FALSE are the same
  saveRDS(model, paste0("processData/",part,"/model_transition", transition_number, ".rds"))
}

# get tidied time-varying data for each transition, and get HR
fit_model <- function(transition, cox_formulas, data_base, data_ms, part){
  ## get tidied time-varying data for each transition
  msdt <- data_ms[trans== transition, ]
  covdt <- data_base[pin%in%msdt$pin,]
  tsdt <- msdt[covdt, on =.(pin = pin, Tstop >= tstop, Tstart <=tstart)][!is.na(trans),]
  # update baseline over time from baseline.
  tsdt <- tsdt[, hypertension_duration := min(Tstart), by = pin][, hypertension_duration:= hypertension_duration/365]
  tsdt[, baseline_yr:=baseline_yr+as.integer(hypertension_duration)]
  tsdt[, baseline_yr_cat := fifelse(baseline_yr <2018, 0, 1)]
  tsdt[, age:=age+as.integer(hypertension_duration)]
  
  tsdt[order(pin, Tstart), `:=`(tstart = Tstart - min(Tstart), tstop = Tstop - min(Tstart)), by = pin]
  tsdt[tstop<time, status := 0]
  tsdt <- as.data.frame(tsdt)
  ## choose formula
  if(transition <=6){formula = cox_formulas$formula1}else{formula = cox_formulas$formula2}

  ## estimate cox model use tsdt for each transition
  cox_analysis(data = tsdt, formula = formula, transition_number = transition, part = part)
}
```


### Step 2: fitted model

```{r}
##################### estimate HR ########################
# define formulas for analysis         
cox_formulas <-list(
  formula1 = Surv(tstart, tstop, status) ~
    exposure + on_adhd_med + sex + baseline_yr_cat + age +
    hypertension_med + t2d + hyperlipid,

  formula2 = Surv(tstart, tstop, status) ~
    exposure + on_adhd_med + sex + baseline_yr_cat + age +
    hypertension_med + t2d + hyperlipid + hypertension_duration) 


for(i in 1:14){
  fit_model(transition = i, cox_formulas = cox_formulas, data_base = dt_base, data_ms = msbmt, part = "adhd")
  print(i)
}
fwrite(dt_base, "processData/adhd/dt_covariate.csv")
fwrite(msbmt, "processData/adhd/dt_ms.csv")
rm(dt)
rm(rx_med)
gc()
```


###Step 3. assess model fitting

```{r}
# cox models at each transition
ls_model <- list()
for(i in 1:14){
  ls_model[[i]] <- readRDS(paste0("processData/adhd/model_transition", i , ".rds"))
  print(i)
  summary(ls_model[[i]]) %>% print()
}
```


# Part 4: Model time-varying cumulative counts of hypertension medication using conditional rate model

```{r}
dt_poisson <- data.frame()

for(i in 1:14){
  ## get tidied time-varying data for each transition
  msdt <- msbmt[trans==i, ]
  covdt <- dt_base[pin%in%msdt$pin,]
  tsdt <- msdt[covdt, on =.(pin = pin, Tstop >= tstop, Tstart <=tstart)][!is.na(trans),]
  # update baseline over time from baseline.
  tsdt <- tsdt[, hypertension_duration := min(Tstart), by = pin][, hypertension_duration:= hypertension_duration/365]
  tsdt[, baseline_yr:=baseline_yr+as.integer(hypertension_duration)]
  tsdt[, baseline_yr_cat := fifelse(baseline_yr <2018, 0, 1)]
  tsdt[, age:=age+as.integer(hypertension_duration)]
  
  tsdt[order(pin, Tstart), `:=`(tstart = Tstart - min(Tstart), tstop = Tstop - min(Tstart)), by = pin]
  tsdt[tstop<time, status := 0]
  tsdt[,days:=tstop-tstart][,counts_hyper_med:=days*hypertension_med]
  tsdt[,cum_hyper_med:=sum(counts_hyper_med), by = pin]
  tsdtp <- tsdt[order(pin, tstart), .SD[1], by = pin][,.(pin, time, cum_hyper_med, trans, exposure, on_adhd_med, sex, age, baseline_yr, baseline_yr_cat, psych_any, t2d, hyperlipid, SES, hypertension_duration)]
  tsdtp <- as.data.frame(tsdtp)
  dt_poisson <- rbind(dt_poisson, tsdtp)
  print(i)
}
setDT(dt_poisson)
dt_poisson[trans <=6, trans_state:="hypertension"]
dt_poisson[trans==7|trans==8, trans_state:="CKD"]
dt_poisson[trans==9|trans==10, trans_state:="stroke"]
dt_poisson[trans==11|trans==12, trans_state:="AMI"]
dt_poisson[trans==13|trans==14, trans_state:="HHF"]
dt_poisson <- dt_poisson[order(pin, trans), .SD[1], by=.(pin, trans)]
dt_poisson <- as.data.frame(dt_poisson)
# Poisson
fit_poisson <- glm(cum_hyper_med ~ exposure + on_adhd_med + sex + baseline_yr_cat + age + t2d + hyperlipid + hypertension_duration + trans_state + offset(log(time)), 
                   data = dt_poisson, family = poisson)

# check poisson model
fit_poisson %>% summary()

# predict rate per day.
newdt <- dt_poisson
newdt$time <-1
pre <- predict(fit_poisson, type="response", newdata = newdt, offset = log(time))
pre %>% summary()

# save poisson model
saveRDS(fit_poisson, "processData/adhd/fit_poisson.rds")
```

